Multiple Sclerosis Diagnosis from Natural Language Processing of Patient Text Data

### 2022-09-12 paper.tex
added some figures

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

@@

\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MSRegister database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms (footnote with list link) from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from ****** diagnosed users in our dataset. \\

@@

\indent The first control group constitutes random users from Twitter and Reddit who have not mentioned the term “MS” or any related terms in their timeline. The dataset was constructed such that there is an equal probability for a potential control user to be selected from Twitter or from Reddit. Moreover, we ignored image-only posts, video-only posts, Twitter retweets and Reddit cross-posts. For Reddit, we added the additional restriction that the potential control group user has never posted on any MS-related subreddit (footnote). After identifying the users, we select random posts in their timeline to be included in our dataset. The number of posts we select per user is controlled such that the final user-comment distribution is kept the same as the diagnosed users in the dataset, as illustrated in Figure *******. Finally, as a measure to prevent posts created by bots to be included in our dataset, no more than 5 duplicate posts were allowed by the same user. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MSRegister dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns (footnote), examples of which can be found in Figure *******. After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

@@

\indent To begin with, we created word clouds from the text given by the MS-diagnosed users (Figure ******a), the generic control one users (Figure ******b), and the diabetes control two users (Figure *****c).  Before creating each of these word clouds, the text was made lowercase, stop words were removed from the text, and the text was tokenized. \\

@@

\indent In addition to generating the word clouds for the dataset, we decided to count the average number of tokens per comment for each part of the dataset. The results of this investigation are shown in Table **********. We observe that the average number of tokens is the greatest for the diabetes control group, lower for the MS-diagnosed control group and lowest for the generic social media control group. Since sentence length varies significantly across the dataset, we need to ensure that the classifiers we train on the dataset do not rely only on the length of the comment for text classification. The implications of this difference are further discussed in the next section.

>>>>

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}

@@

\begin{figure}[h!]
  \centering
   \includegraphics[width=\linewidth]{usercommentdistribution.png}
   \caption{User-Comment Distribution}
  \label{fig:usercomment}
\end{figure}
\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MSRegister database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms (footnote with list link) from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from 3037 diagnosed users in our dataset. \\

@@

\indent The first control group constitutes random users from Twitter and Reddit who have not mentioned the term “MS” or any related terms in their timeline. The dataset was constructed such that there is an equal probability for a potential control user to be selected from Twitter or from Reddit. Moreover, we ignored image-only posts, video-only posts, Twitter retweets and Reddit cross-posts. For Reddit, we added the additional restriction that the potential control group user has never posted on any MS-related subreddit (footnote). After identifying the users, we select random posts in their timeline to be included in our dataset. The number of posts we select per user is controlled such that the final user-comment distribution is kept the same as the diagnosed users in the dataset, as illustrated in Figure \ref{fig:usercomment}. Finally, as a measure to prevent posts created by bots to be included in our dataset, no more than 5 duplicate posts were allowed by the same user. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MSRegister dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns (footnote), examples of which can be found in Figure ****. After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

\begin{figure*}[h!]
  \centering
  \begin{subfigure}[b]{0.3\linewidth}
    \includegraphics[width=\linewidth]{msregisterfinal.png}
    \caption{Diagnosed Users (MS)}
  \end{subfigure}
  \hspace{1em}
  \begin{subfigure}[b]{0.3\linewidth}
    \includegraphics[width=\linewidth]{controlonefinal.png}
    \caption{Control One (Generic)}
  \end{subfigure}
  \hspace{1em}
  \begin{subfigure}[b]{0.3\linewidth}
    \includegraphics[width=\linewidth]{controltwofinal.png}
    \caption{Control Two (Diabetes)}
  \end{subfigure}
  \caption{Word Clouds For Different Parts Of The Dataset}
  \label{fig:wordclouds}
\end{figure*}

@@

\indent To begin with, we created word clouds from the text given by the MS-diagnosed users (Figure \ref{fig:wordclouds}a), the generic control one users (Figure \ref{fig:wordclouds}b), and the diabetes control two users (Figure \ref{fig:wordclouds}c).  Before creating each of these word clouds, the text was made lowercase, stop words were removed from the text, and the text was tokenized. \\

@@

\indent In addition to generating the word clouds for the dataset, we decided to count the average number of tokens per comment for each part of the dataset. The results of this investigation are shown in Table ****. We observe that the average number of tokens is the greatest for the diabetes control group, lower for the MS-diagnosed control group and lowest for the generic social media control group. Since sentence length varies significantly across the dataset, we need to ensure that the classifiers we train on the dataset do not rely only on the length of the comment for text classification. The implications of this difference are further discussed in the next section.
\begin{table}
\centering
\begin{tabular}{p{0.5\linewidth}p{0.4\linewidth}}
\hline
\textbf{Dataset} & \textbf{Average number of tokens per comment}\\
\hline
Diagnosed (MS) & {29.1} \\
Control One (Generic) & {14.1} \\
Control Two (Diabetes) & {85.3} \\ 
\hline
\end{tabular}
\caption{Average number of tokens per comment for evry part of the dataset}
\label{tab:}
\end{table}
